---
title: "Arithmetic in R"
output:
 pdf_document:
   toc: yes
   toc_depth: 4
  html_notebook:
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
    theme: darkly
---
# Arithmetic in R

As you've seen in the live coding, R can be used to solve equations and print the results.  This next code chunk demonstrates simple math, a variable assignment and 2 common ways to print the value stored in a variable.

```{r}
# print the result of a mathematical equation directly, add the integers 22 and 7

```

```{r}
# store the result in a variable named x.

```

```{r}
# Execute this block to see 2 different ways to print the value stored in x:
# x
# print(x)
```

We've assigned an object to the variable x, which now resides in our global environment.  In the pane in the top right, if you click the "Environment" you can see your global variables.  At this point, lets clear our environment to keep things tidy.  As we progress through this notebook, keep your eye on the environment to inspect your variables without having to print them out.

```{r}
# Execute this block to clear all objects in our environment
rm(list=ls())
```

## Golden Ratio $\phi$ (phi)

The golden ratio is a special irrational number (approximately equal to 1.618) that appears throughout mathematics, geometry, art, architecture, and other areas.  It is defined by the following equation:

$\phi = \frac{(1 + \sqrt 5)}{2}$

Using basic arithmetic in R, compute the golden ratio and store that value in a variable called **phi** and then print phi.  Recall that to find the square-root of a number, you can raise that number to the power of 1/2, to find the cube-root, raise that number by 1/3, etc.  Remember that in R, the assignment operator is **<-**, for example: **var_name <- 3**.

```{r}
# compute and print the golden ratio

```

## Seven different approximations of $\pi$ (pi)

$\pi$ is an irrational number defined as the ratio of a circle's circumference to its diameter.

1. $\pi = \frac{22}{7}$

2. $\pi = \sqrt 2 + \sqrt 3$

3. $\pi = \sqrt[3]{31}$

4. $\pi = \sqrt{7 + \sqrt{6 + \sqrt{5}}}$

5. $\pi = \frac{355}{113}$

6. $\pi = \sqrt[4]{\frac{2143}{22}}$

7. $\pi = \frac{80\sqrt{15}(5^4 + 53\sqrt{89})^\frac{3}{2}}{3308(5^4 + 53\sqrt{89}) - 3\sqrt{89}}$

Compute these 7 different approximations of $\pi$ where the value of the first approximation is stored in a variable named **pi.1**, the second in **pi.2**, and so forth, then print these variables.

```{r}
# compute and print the 7 approximations of pi

```

R has a built-in $\pi$ constant, called **pi**.  Run this next code chunk to check out the value of this constant.

```{r}
# execute this block to see the value of the built-in constant
pi
```

Here we show how to use the sprintf() function to print the floating-point number $\pi$ (R constant pi) with 25 digits of precision.  Click on the sprintf() function below and hit **F1** on your keyboard to get help on this function.

```{r}
# R wraps the C sprintf() function for string formatting.  Execute this block to see the output.
sprintf("%.25f", pi)
```

In the next chunk, use sprintf() to print the floating point number $\phi$ (which you calculated in the first code chunk) with 50 digits of precision.

```{r}
# use sprintf() to see the value of pi with 50 digits of precision

```

In the **Examples** section of the sprintf() help (near the bottom), *copy-and-paste the "various formats of pi" from the help text into the next code chunk*, run it and note the various output formats.

```{r}

```

### Assignment differences between = and <- operators

Here is an example context where the = and <- operators behave differently.

```{r}
# execute this block and observe what happened
median(x = 1:10)
# This next line will fail
#x
```

```{r}
median(x <- 1:10)
# This next line will succeed
# x
```

Congratulations!  You've learned how to open an R notebook in RStudio, do basic mathematical calculations, variable assignments, insecting variables in the global environment, cleaning variables in the global environment, output formatting using the sprint() function, code comments, how to get help on functions, and how to knit the R notebook into an HTML document for publishing.

Also, be sure to take note of some of the basics of Rmarkdown in this document, including:

- Title
- HTML output format (commented is PDF output format (requires pdflatex))
- Table of contents
- Sections
- Numbered lists
- Bulleted lists
- Math notations
- Italic text
- Bold text
- Code chunks

Quite a bit was learned in this lab.  If you still have time, feel free to experiment farther, for example, calculate the differences between the seven approximations of $\pi$ and the built-in R constant and print the results to 50 digits of precision.